Canary in the Coal Mine: Anti-NGO Legislation as a Predictor of Civil Society Repression

### 2021-03-27 manuscript/manuscript.Rmd
Suparna additions

Governments across the globe have been cracking down on civil society organizations (CSOs) -- a phenomenon referred to as closing civic space in policy arena. The non-violent nature of this repression is particularly notable.The last two decades have
seen a proliferation of laws designed to limit the influence of civil society organizations,
with more than 100 countries obstructing, repressing, or closing the legal
environment for civil society have required that funds
to NGOs be routed through state-owned banks, mandated that funds cannot be used
for political purposes, or prohibited NGOs from accepting any funds from foreign
sources (Carothers, DPR, Chaudhry, Heiss)

As someone points out, states have toned down 'open, violent confrontation' and favor the development of a 'civilized' oppression. The use of legal tools instead of violent repression means that NGOs have to  their practices to stay within the bounds of permitted activities and operations.  This has led to depoliticization, avoidance of direct political activism. But that state cannot soften or neutralize all forms of political activities in subtle, legal ways that they have with NGOs. Soon or later, states need to deal with threats in a more violent, confrontational manner. Studying these NGO laws can thus be the canary in the coal mine which signals the states that intend to continue down this repressive path, especially in their short to medium term trajectories.

What does their targeting tell us about a state's broader respect for civil society and the overall level of human rights repression? Can the crackdown on NGOs act as an indicator for subsequent increasing repression?

Information about crackdown can be a valuable indicator of human rights trajectories. Can act as an early warning signal for worsening repression, suggesting which countries are testing the waters with comparatively 'milder' forms of repression before enacting more violent forms of crackdown.

Structural factors such as levels of democracy, economic development, electoral competitiveness can identify countries that are likely to have more or less respect for civic freedoms, but these factors are also slow to change over time. Thus, they may not be particularly suitable for forecasting short-term trajectories, especially for moderatively repressive countries.

@@

Why we look as NGOs
From an economic perspective, the foreign aid community prefers NGOs since
they overcome many failures associated with channeling funds through the state
(Dietrich 2013). They provide services to large parts of the population in the Global
South, especially where governments are unable to do so on a consistent basis.

## Structural factors explaining repression

## Exploring non-violent forms of repression
Why might repressive states start with non-violent forms of repression?

# State repression of NGOs

A robust, strong civil society is widely considered as an important component on democratic transition and consolidation.
Accountability
Corruption
Monitoring
In addition, services when the state falls short
Leaders want to avoid accountability. Shape the narrative of their performance both domestically and internationally. Don't like being shamed.
Can be the beginning of a deteriorating human rights situation.

@@

>>>>

Governments across the globe have been cracking down on civil society organizations (CSOs) -- a phenomenon referred to as closing civic space or shrinking civic space (CITE). The non-violent nature of this repression is particularly notable. The last two decades have seen a proliferation of laws designed to limit the influence of civil society organizations, with more than 100 countries obstructing, repressing, or closing the legal environment for civil society. Since 2013, 91 countries had proposed or enacted 244 measures restricting civil society [@icnl:2021]. In Human Rights Watch's 2016 World Report, Executive Director Kenneth Roth argued that civil society was under more aggressive attack than at any time in recent memory, and over the past decade, the majority of the world has seen a substantial narrowing of allowable civic space. As seen in Figure \@ref(fig:civicus-map), by the end of 2020, only `r civicus$Open$prop_nice` of countries had open and unrestricted civil societies [@civicusmonitor; @ChaudhryHeiss:2021].

WHAT KINDS OF LAWS, WHAT DO THESE LAWS DO.

The use of legal tools instead of violent repression means that NGOs have to adapt their practices to stay within the bounds of permitted activities and operations. While in some cases this has led to the depoliticization of NGOs and avoidance of direct political activism by CSOs, states cannot soften or neutralize all forms of political activities in subtle, legal ways that they have with NGOs. Soon or later, states need to deal with threats in a more violent, confrontational manner. Studying these NGO laws can thus be the canary in the coal mine which signals the states that intend to continue down this repressive path. What does state targeting of NGOs tell us about a state's broader respect for civil society and the overall level of human rights repression? Can the crackdown on NGOs act as an indicator for subsequent increasing repression?

ARGUMENT, METHODS, AND FINDINGS. 
In this paper, we argue that information about NGO crackdown can be a valuable indicator of human rights trajectories. Such crackdown can act as an early warning signal for worsening repression, suggesting which countries are testing the waters with comparatively 'milder' forms of repression before enacting more violent forms of crackdown. To assess if attacks on NGOs and activists can provide us useful information about future repression, we use an original dataset on state crackdown on NGOs using legal repression across all countries from 1990-2013. WHAT ELSE DO WE USE. Our analysis shows FILL

Our paper makes an important contribution in predicting (IS THIS THE RIGHT WORD) the onset of repressive state policies. While structural factors such as levels of democracy, economic development, electoral competitiveness can identify countries that are likely to have more or less respect for civic freedoms, these factors are also slow to change over time. As such, they may not be particularly suitable for forecasting short-term trajectories, especially for moderatively repressive countries. On the other hand, state repression of NGOs is more dynamic and provides information about changing levels of human rights standards and repression in subsequent years.

ROADMAP

@@

Repression is the use of coercive action against an individual or organization, within the territorial jurisdiction of the state, for the purpose of imposing a cost on the target as well as deterring specific activities and/or beliefs perceived to be challenging to government personnel, practices, or institutions (Goldstein 1978, xxvii; Poe and Tate 1994; Davenport 2007b; and Ritter and Conrad 2016).

One part of the existing literature focuses on the role of structural factors such as levels of democracy, economic development, electoral competitiveness in predisposing a state to increased use of repression. Most notably, both democracies and democratization is found to have a pacifying effect on state repression (Davenport and Armstrong 2004, Krain 1997, Richards 1999). Democracies are found to have a negative effect on repression largely because of the increased probability of sanctions against authorities for undesirable behavior (Davenport 2007, 41).

Another strand of research focuses on how the behavior of various actors shapes government response. One of the most consistent findings in the political violence literature is that authorities repress to counter ongoing challenges to the status quo. In other words, dissent increases repressive behavior (Davenport 2007).  However, dissent and repression have an endogenous relationship - state authorities also frequently engage in preventive repression to undermine or restrict groups’ abilities to dissent (Ritter and Conrad 2016; and Conrad and Ritter 2019.) In addition to attacking groups’ willingness to organize against states, such
preventive repression also damages their capacity to impose costs (Davenport 2007a, 47.)

NGOs are one actor that can pose challenges to political authorities, and state repression of these actors, especially through legal restrictions or administrative crackdown can also be considered a form of preventive repression. Below, we explore why government repress NGOs, especially through subtle, legal means.

## State repression of NGOs

A robust, strong civil society is widely considered as an important component on democratic transition and consolidation (Putnam). Politically, as important civil society actors, NGOs enable citizens to voice their grievances, seek redress, and carry out lobbying and advocacy activities to further their interests. They help keep government accountable, monitor human rights violations and corruption. From an economic perspective, NGOs provide services when the state falls short.

As a result of these roles, NGOs can wield real influence, which can be threatening to states. State crackdown on NGOs has become pervasive as these organizations have the ability to influence electoral politics (Dupuy, Ron, and Prakash 2016, Weiss 2009), aid mobilization (Murdie and Bhasin 2011; Bell et al. 2013; and Boulding 2014), and threaten a state’s economic interests (Dietrich and Murdie 2017; Lebovic and Voeten 2009; and Nielsen 2013.).

Why might repressive states start with non-violent forms of repression?
WHY THE GROWTH IN THESE LAWS
As someone points out, states have toned down 'open, violent confrontation' and favor the development of a 'civilized' oppression.

Leaders want to avoid accountability. Shape the narrative of their performance both domestically and internationally. Don't like being shamed.
Can be the beginning of a deteriorating human rights situation.

@@

##Expectations

#Methods

#Results and Analysis

#Conclusion and Implications
